### 2017-07-26 chapters/introduction/patient.tex
Write summary. Only missing to write about fenics and summary of  the unpublished manuscripts.

@@

\paragraph{The volume observation operator
  $\mathcal{H}_{\mathrm{volume}}$}
In some cases, the ventricular cavity volume can easily be computed
analytically from the displacement field $\uvec$. Denote the inner
cavity of the left ventricle in the current configuration by
$\omega_{\mathrm{endo}}$. Further let $\mathrm{d} v$ and $\xvec$
denote an infinitesimal volume element and the coordinate in the
current configuration respectively. Then by the divergence theorem we have
\begin{align}
  \mathcal{H}_{\mathrm{volume}}(\uvec)= \int_{\omega_{\mathrm{endo}}} \mathrm{d} v =
  \frac{1}{3}\int_{\omega_{\mathrm{endo}}} \nabla \cdot \xvec \mathrm{d} v =
  \frac{1}{3}\int_{\partial \omega_{\mathrm{endo}}} \xvec \cdot \mathbf{n} \mathrm{d} s, 
\end{align}
where $\mathbf{n}$ and $\mathrm{d} s$ are the unit normal and a surface
element on the current configuration respectively.
Note that the boundary $\partial \omega_{\mathrm{endo}}$ includes the
endocardial basal boundary. However in the case when the base is flat
and fixed in longitudinal direction, the contribution from this
boundary integral will be zero. In this case the boundary integral
will be the same as the integral over the endocardial boundary, with
the only change being the change of sign on the normal vector. By
Nansons formula we obtain 
\begin{align}
  \mathcal{H}_{\mathrm{volume}}(\uvec)= -\frac{1}{3}\int_{\lvendo} \left( \Xvec + \uvec \right) J \F^{-T} \Nvec \mathrm{d}S,
  \label{eq:volume_operator}
\end{align}
where now $\lvendo$,  $\Xvec$, $\Nvec$ and $\mathrm{d}S$ are respectively the
endocardial surface, the reference coordinate, the unit normal and a
surface element in the reference configuration.

@@

We could instead plug in the solution to
\eqref{eq:tangent_linear_model} to get
\begin{align}
   D  \widehat{\mathcal{J}}
    % &=  D_{\mu}\mathcal{J}(\state(\mu), \mu)
      % - D_{\state}\mathcal{J}(\state(\mu), \mu) \left(  D_{\state} \mathcal{M}(\state(\mu), \mu) \right)^{-1} D_{\mu} \mathcal{M}(\state(\mu), \mu) \\
      &= D_{\mu}\mathcal{J}(\state(\mu), \mu) - z^* D_{\mu} \mathcal{M}(\state(\mu), \mu),
        \label{eq:functional_gradient}
\end{align}
where
\begin{align*}
  z^* = D_{\state}\mathcal{J}(\state(\mu), \mu) \left(  D_{\state} \mathcal{M}(\state(\mu), \mu) \right)^{-1} \\ 
  \implies z^* D_{\state} \mathcal{M}(\state(\mu), \mu) =  D_{\state}\mathcal{J}(\state(\mu), \mu)
  \label{eq:adjoint_equation}
\end{align*}
Note that in \eqref{eq:functional_gradient}, the variable $z^*$
represents a Lagrange multiplier enforcing the PDE constraint.  
In order to solve this system we need to have on the form $A\xvec =
b$, and we take the adjoint (Hermitian transpose) to get
\begin{align}
  D_{\state} \mathcal{M}(\state(\mu), \mu)^* z = D_{\state}\mathcal{J}(\state(\mu), \mu)^*
\end{align}
Equation \eqref{eq:adjoint_equation} is called the adjoint equation.
To compute the gradient we can now solve for $z$ and then insert $z^*$
into \eqref{eq:functional_gradient} to compute the gradient.

@@

\subsubsection{Multiobjetive optimization}
Parameters in the model are estimated based on minimizing a cost
functional representing the mismatch between observations and
simulations. Assume we are given $N$ different observations, then in
principle we have $N$ different cost functionals to minimize:
\begin{equation}
  \begin{aligned}
    \label{eq:multiobjetive_opt}
    & \underset{\mu \in P}{\text{minimize}}
    & &  \left\{ \mathcal{J}_1(\state, \mu), \mathcal{J}_2(\state, \mu), \cdots, \mathcal{J}_N(\state, \mu) \right\} \\
    & \text{subject to}
    & &\mathcal{M}(\state, \lambda) = 0
  \end{aligned}
\end{equation}
with
\begin{align}
  \mathcal{J}_i(\mu) = \left( \mathbf{y} - \mathcal{H}(\state(\lambda)) \right)^2, \; i = 1,\cdots, N
\end{align}
The above observations comes from the same source, and hence it should
be intuitive that minimizing one of them, would also bring the other
closer to a minimum. However, with the presence of noise in the data,
this might not always be the case, and hence we have conflicting
objectives. Problem \eqref{eq:multiobjetive_opt} is called a
multiobjetive optimization problem \cite{deb2016multi}. Such problems
typically do not have a single optimal solution, but a family of
solutions called \emph{Pareto optimal solutions}. The basic methods for
solving such problems includes \emph{the weighted method} and \emph{the
  $\varepsilon$-constraint method}. In this thesis we have only used
the weighted method, which is based on minimizing a weighted sum of
the individual cost functionals, i.e:

@@

\subsubsection{Regularization}
Regularization is an important concept in many scientific disciplines
concerned with data fitting. When using data that essentially are
realizations of a stochastic process, we need to take into account
that the data are corrupted with noise. Moreover, the model we are using
is a simplification of a physical process, and fitting the
data exactly to the model is unrealistic. This is especially
important if the number of parameters we are trying to estimate are
more than the number of data points used as input. In such cases it is often
beneficial to make some assumptions about the solution we are looking
for in order to avoid overfitting. For example, we may assume that the
parameter we are searching for is smooth. Suppose that $\mu \in
H^1(\Omega)$ is the parameter we want to find, and that we want to
favor smooth solutions. Then the norm of the gradient $\| \nabla \mu
\|$ should be small. One way to achieve this is to add a penalty term
to the objective functional so that in stead of minimizing
$\mathcal{J}(\state, \mu)$ we minimize $\mathcal{J}(\state, \mu) +
\lambda \| \nabla \mu\|$, where the constant $\lambda$, known as the
regularization parameter, controls the amount of regularization.
Note that in the limit $\lambda \rightarrow \infty$ we would end up
with a constant function. This approach is referred to as Tikhonov
regularization.

@@

\subsection{Identifiability of parameters}
% Try dividing heart mesh is different segment, discuss uniqueness of
% paraperers
% Discuss regularization technices
When dealing with parameter estimation, you should always consider
questions about identifiability and uniqueness. This depends on the
model, the data and the objective functional
\cite{hadjicharalambous2015analysis}. A first test should be to
generate synthetic data with some prescribed parameters, feed this
error-free data to the data assimilation method, and ensure that you
are able to retrieve the parameters that generated that data. If you
end up with a different parameter set than the one generated the data,
your problem is not \emph{structurally identifiable}
\cite{chabiniok2016multiphysics}. Your problem is said to be
\emph{practical identifiable} if the parameters can be determined
uniquely based on the data at hand. If the test passes you should try
to add noise to the data, and see if your data assimilation is stable
with respect to noise in the measurement.

@@

\subsection{Mechanical biomarkers}
As can be seen from Figure \ref{fig:cardiac_imaging_model}, data
assimilation of computational models and medical imaging can be used to
extract biomarkers which potentially can be used to gain further
insight into ...

>>>>

\subsubsection{Coupled material parameter estimation and unloading
  algorithm}
\label{sec:intro_coupled_material}
One thing to note is that the unloaded configuration depends upon the
chosen material parameters, which are generally also unknown. For
example, applying the backward displacement method to a geometry will
result in a smaller unloaded cavity volume if the material is softer.
Therefore, the material parameters estimation is coupled to the
problem of fined the unloaded, zero pressure geometry.
In \cite{nikou2016effects}, Nikou et al propose a method for estimating
material parameters as well the unloaded geometry by iterating between
the two methods, and terminate if the difference in unloaded cavity
volume is less than $5 \%$. This method is general and could be applied
using any unloading algorithm and any method to estimate the material
parameters.

Assume we are given an image-based LV geometry (which is not
end-diastole) with measured pressure $p_{\mathrm{LV}}$. Further assume
that we are given the pressure and  volume at end-diastole,
$p_{\mathrm{LV}}^{\mathrm{ED}}$ and $V_{\mathrm{LV}}^{\mathrm{ED}}$
respectively. Then we want to find
\begin{enumerate}
  \item \emph{the unloaded (zero pressure) geometry} so that when loaded with
    $p_{\mathrm{LV}}$ gives back the image based geometry, and
   \item \emph{material parameters}, so that when we load the
     estimated unloaded geometry with $p_{\mathrm{LV}}^{\mathrm{ED}}$
     it has a cavity volume equal to $V_{\mathrm{LV}}^{\mathrm{ED}}$.     
\end{enumerate}

The iterative strategy is presented in Algorithm \ref{alg:intro_unload_mat}.
\begin{algorithm}
  \caption{Uloaded material parameter estimation\label{alg:intro_unload_mat}}

\begin{algorithmic}[1]

  \State Initialize $\mathbf{m}_0$ \Comment{intial guess for material parameters}
  \State $i = 0$

  \While{difference in reference cavity volumes $>5 \%$ and $i < 15$}

  \State \parbox[t]{\dimexpr\linewidth-\algorithmicindent}{Find the
    unloaded geometry using any of the above methods and using
    $\mathbf{m}_{i}$ as material parameters.\strut} 
  \State \parbox[t]{\dimexpr\linewidth-\algorithmicindent}{Estimate
    material parameters $\mathbf{m}_{i+1}$ so that when we load the
    estimated unloaded geometry with $p_{\mathrm{LV}}^{\mathrm{ED}}$
    it has a cavity volume equal to $V_{\mathrm{LV}}^{\mathrm{ED}}$. \strut}    
  \State $i = i + 1$
  \EndWhile

\end{algorithmic}
\end{algorithm}

Note that this method apply also to bi-ventricular geometries and to
cases when more measurement points are available for estimating the
material parameters.

@@

\paragraph{The volume observation operator
  $\mathcal{H}_{\mathrm{volume}}$}
In some cases, the ventricular cavity volume can easily be computed
analytically from the displacement field $\uvec$. Denote the inner
cavity of the left ventricle in the current configuration by
$\omega_{\mathrm{endo}}$. Further let $\mathrm{d} v$ and $\xvec$
denote an infinitesimal volume element and the coordinate in the
current configuration respectively. Then by the divergence theorem we have
\begin{align}
  \mathcal{H}_{\mathrm{volume}}(\uvec)= \int_{\omega_{\mathrm{endo}}} \mathrm{d} v =
  \frac{1}{3}\int_{\omega_{\mathrm{endo}}} \nabla \cdot \xvec \mathrm{d} v =
  \frac{1}{3}\int_{\partial \omega_{\mathrm{endo}}} \xvec \cdot \mathbf{n} \mathrm{d} s, 
\end{align}
where $\mathbf{n}$ and $\mathrm{d} s$ are the unit normal and a surface
element on the current configuration respectively.
Note that the boundary $\partial \omega_{\mathrm{endo}}$ includes the
endocardial basal boundary. However in the case when the base is flat
and fixed in longitudinal direction and located at the $x=0$ plane,
the contribution from this boundary integral will be zero. In this case the boundary integral
will be the same as the integral over the endocardial boundary, with
the only change being the change of sign on the normal vector. By
Nansons formula we obtain 
\begin{align}
  \mathcal{H}_{\mathrm{volume}}(\uvec)= -\frac{1}{3}\int_{\lvendo} \left( \Xvec + \uvec \right) J \F^{-T} \Nvec \mathrm{d}S,
  \label{eq:volume_operator}
\end{align}
where now $\lvendo$,  $\Xvec$, $\Nvec$ and $\mathrm{d}S$ are respectively the
endocardial surface, the reference coordinate, the unit normal and a
surface element in the reference configuration.

@@

We could instead plug in the solution to
\eqref{eq:tangent_linear_model} to get
\begin{align}
   D  \widehat{\mathcal{J}}
    % &=  D_{\mu}\mathcal{J}(\state(\mu), \mu)
      % - D_{\state}\mathcal{J}(\state(\mu), \mu) \left(  D_{\state} \mathcal{M}(\state(\mu), \mu) \right)^{-1} D_{\mu} \mathcal{M}(\state(\mu), \mu) \\
      &= D_{\mu}\mathcal{J}(\state(\mu), \mu) - z^* D_{\mu} \mathcal{M}(\state(\mu), \mu),
        \label{eq:functional_gradient}
\end{align}
where
\begin{align*}
  z^* = D_{\state}\mathcal{J}(\state(\mu), \mu) \left(  D_{\state} \mathcal{M}(\state(\mu), \mu) \right)^{-1} \\ 
  \implies z^* D_{\state} \mathcal{M}(\state(\mu), \mu) =  D_{\state}\mathcal{J}(\state(\mu), \mu)
  \label{eq:adjoint_equation}
\end{align*}
Note that in \eqref{eq:functional_gradient}, the variable $z^*$
represents a Lagrange multiplier enforcing the PDE constraint.  
In order to solve this system, we need to have it on the form $A\xvec =
b$, and we take the adjoint (Hermitian transpose) to get
\begin{align}
  D_{\state} \mathcal{M}(\state(\mu), \mu)^* z = D_{\state}\mathcal{J}(\state(\mu), \mu)^*
\end{align}
Equation \eqref{eq:adjoint_equation} is called the adjoint equation.
To compute the gradient we can now solve for $z$ and then insert $z^*$
into \eqref{eq:functional_gradient} to compute the gradient.

@@

\subsubsection{Multiobjetive optimization}
Parameters in the model are estimated based on minimizing a cost
functional representing the mismatch between observations and
simulations. Assume we are given $N$ different observations, then in
principle we have $N$ different cost functionals to minimize:
\begin{equation}
  \begin{aligned}
    \label{eq:multiobjetive_opt}
    & \underset{\mu \in P}{\text{minimize}}
    & &  \left\{ \mathcal{J}_1(\state, \mu), \mathcal{J}_2(\state, \mu), \cdots, \mathcal{J}_N(\state, \mu) \right\} \\
    & \text{subject to}
    & &\mathcal{M}(\state, \mu) = 0
  \end{aligned}
\end{equation}
with
\begin{align}
  \mathcal{J}_i(\mu) = \left( \mathbf{y} - \mathcal{H}(\state(\mu)) \right)^2, \; i = 1,\cdots, N
\end{align}
The above observations comes from the same source, and hence it should
be intuitive that minimizing one of them, would also bring the other
closer to a minimum. However, with the presence of noise in the data,
this might not always be the case, and hence we have conflicting
objectives. Problem \eqref{eq:multiobjetive_opt} is called a
multiobjetive optimization problem \cite{deb2016multi}. Such problems
typically do not have a single optimal solution, but a family of
solutions called \emph{Pareto optimal solutions}. The basic methods for
solving such problems includes \emph{the weighted method} and \emph{the
  $\varepsilon$-constraint method}. In this thesis we have only used
the weighted method, which is based on minimizing a weighted sum of
the individual cost functionals, i.e:

@@

\subsubsection{Regularization}
\label{sec:intro_regularization}
Regularization is an important concept in many scientific disciplines
concerned with data fitting. When using data that essentially are
realizations of a stochastic process, we need to take into account
that the data are corrupted with noise. Moreover, the model we are using
is a simplification of a physical process, and fitting the
data exactly to the model is unrealistic. This is especially
important if the number of parameters we are trying to estimate are
more than the number of data points used as input. In such cases it is often
beneficial to make some assumptions about the solution we are looking
for in order to avoid overfitting. For example, we may assume that the
parameter we are searching for is smooth. Suppose that $\mu \in
H^1(\Omega)$ is the parameter we want to find, and that we want to
favor smooth solutions. Then the norm of the gradient $\| \nabla \mu
\|$ should be small. One way to achieve this is to add a penalty term
to the objective functional so that in stead of minimizing
$\mathcal{J}(\state, \mu)$ we minimize $\mathcal{J}(\state, \mu) +
\lambda \| \nabla \mu\|$, where the constant $\lambda$, known as the
regularization parameter, controls the smoothness of $\mu$.
Note that in the limit $\lambda \rightarrow \infty$ we would end up
with a constant value of $\mu$. This approach is referred to as Tikhonov
regularization.

@@

\subsection{Identifiability of parameters}
\label{sec:intro_identifiability}
% Try dividing heart mesh is different segment, discuss uniqueness of
% paraperers
% Discuss regularization technices
When dealing with parameter estimation, you should always consider
questions about identifiability and uniqueness. This depends on the
model, the data and the objective functional
\cite{hadjicharalambous2015analysis}. A first test should be to
generate synthetic data with some prescribed parameters, feed this
error-free data to the data assimilation method, and ensure that you
are able to retrieve the parameters that generated that data. If you
end up with a different parameter set than the one generated the data,
your problem is not \emph{structurally identifiable}
\cite{chabiniok2016multiphysics}. Your problem is said to be
\emph{practical identifiable} if the parameters can be determined
uniquely based on the data at hand. If the test passes you should try
to add noise to the data, and see if your data assimilation is stable
with respect to noise in the measurement.

@@

\subsection{Mechanical biomarkers}
As can be seen from Figure \ref{fig:cardiac_imaging_model}, data
assimilation of computational models and medical imaging can be used to
extract biomarkers. By biomarker, we mean an indicator of some
biological of physiological process that give insight into the state
or condition of that process. Examples of biomarkers that can be
extracted from a computational models are parameters related to tissue
stiffness, like the parameters in the Holzapfel-Ogden material law
\eqref{eq:holzapel_trans}, or parameters related to contractility like
the active stress ($T_a$ in \eqref{eq:intro_active_stress}) or the
active strain ($\gamma$ in \eqref{eq:intro_active_strain_Fa_gjerald}).
Computational models opens a new door when it comes to extraction of
biomarkers. Mechanical features such as stress, contractility and
elastance can easily be computed using a computational model. Of
course, validation of these models is what matter most in order to
make these biomarkers clinically useful.

% Extracting biomarkers from computational models can be used
% to gain insight into physiological processes by e.g see how
% physiology changes when parameters or input are perturbed
% \cite{remme2011mechanics}. A model allows you to test

% which potentially can be used to gain further
% insight into

### 2017-07-26 chapters/introduction/summary.tex
Add latest
Write summary. Only missing to write about fenics and summary of  the unpublished manuscripts.

\section{Summary of papers}

\subsection{Paper 1}

\subsection{Paper 2}

\subsection{Paper 3}

\subsection{Paper 4}

\section{Other contributions}
Along with the research articles presented in this thesis, other
types of contributions in terms of talks, posters and software has
been made during the writing of this thesis. These contributions are
listed below.

@@

\subsection{Software}
  \urlstyle{rm}
\begin{itemize}
  \item Pulse-Adjoint, FEniCS-based cardiac mechanics solver and data
    assimilator, source: \url{https://bitbucket.org/finsberg/pulse_adjoint}
  \item Mesh-Toolbox, Toolbox for generating FEniCS meshes from 4D
    Echo,  source: \url{https://bitbucket.org/finsberg/mesh_generation}
\end{itemize}

\section{Closing remarks and future directions}

A couple remarks regarding the limitations of the methods used in this
thesis shoud be made. Here we start by listing a couple of statements
which should be taken into considerations.

\begin{itemize}
\item \emph{The quality of features/biomarkers you can extract from a data-driven
model cannot be any better than the data used as input to the
model}. Most of the results presented in this thesis are based on data
obtained from clinical measurements of real patients.
\item \emph{The spatial resolution of the parameters should be
    reflected in the spatial resolution of the observations.} When
  trying to fit data that are spatially resolved at some level,
  choosing parameters that are resolved at a finer level should be
  done with caution. The continuity in the underlying physics as well
  as regularization techniques could be used to ...

\end{itemize}

During the work of this thesis, several questions still remains open
and would require

\begin{itemize}
\item The active model for the myocardium is has .. Degree of
  tranverse activation. Matching of strain data..
\item Identifiability of parameters... Uniqueness of
  solutions.. Amount of regularization.. Convexity of the mismatch functional
\item Appropriate boundary conditions.. In paper three we saw big
  differences in the choice of boundary conditions. Especially, the
  magnitude of the stress seems to 
\item Coupling of electrophysiologigy and mechanics in an
  ajoint-based. data assimilation framework. 
\end{itemize}

>>>>

\section{Summary of papers}
In this final introductory section we summarize the papers that make
up this thesis, and we also give some final concluding remarks
and future directions.

\subsection{Paper 1: High-resolution data assimilation of cardiac mechanics
  applied to a dyssynchronous ventricle}
In this paper we develop and test the pipeline for constructing a
patient specific mechanical simulation of a patient's heart, based on
clinical measurments and adjoint-based data assimilation
techniques.

As a model case we consider one patient, diagnosed with
left bundle branch block and selected for cardiac resynchronization
therapy (CRT). Prior to the CRT implantation the patient had 4D
echocardiography taken for which the LV geometry, LV volumes and LV
regional strains throughout the cardiac cycle are measured. During
implantation of the CRT device the LV pressure where also measured
invasively. The LV pressure measurements are used as boundary condition at the
endocardium, while the LV volume and LV regional strain is incorporated
into a cost functional that we want to minimize.

The pipeline is divided into two phases, a passive phase where we
estimate the linear isotropic parameters $a$ in
\eqref{eq:holzapel_trans} as a global material parameter using the
measurement points belonging atrial systole, and a active phase where we estimate the
a spatially varying contraction parameter ($\gamma$ in
\eqref{eq:intro_active_strain_Fa_gjerald}) at each measurement point
with active contraction. During the passive phase we only fit the
volumes, while during the active phase 51 strain measurements in the radial,
longitudinal and circumferential direction in each AHA segment (Figure
\ref{fig:echopac_output}) are used in the optimization.

The results show an excellent fit with measured strain an volume, with
a average relative error in the volume and strain of less than 0.4 \% and 3 \%
respectively using a contraction parameter with one degree of freedom
for each vertex in the geometry (2661 parameters in total). Parameters
at lower spatial resolution are also tested to show the necessity of
high spatial resolution to fit the measured strain data.
A synthetic test is also performed in order to show that that method
is able to recapitulate the generated data, also with noise added to
the data. Moreover, a sensitivity analysis to different parameters are
presented in the appendix.

\subsection{Paper 2: Estimating cardiac contraction through high resolution
  data assimilation of a personalized mechanical model}
In this paper we apply the method developed in the previous paper to a
cohort of patients and estimate indices of cardiac contractility. More
specifically, a group of seven patients diagnosed with left bundle branch block and
selected for cardiac resynchronization therapy (CRT) and a group of
seven healthy control subjects where included in the study.

The pipeline is similar to the one outlined in paper 1, with the
exception that we also estimate the unloaded, stress-free
configuration using the method outlined in Section
\ref{sec:intro_coupled_material}.

The optimized active strain parameter in
\eqref{eq:intro_active_strain_Fa_gjerald} as well as the optimized
active stress parameter in \eqref{eq:intro_active_stress} are averaged
over the ventricle and compared between the two groups. The healthy
group showed a significant increase in both of these
parameters. Furthermore, estimation of end-systolic elastance by
perturbation of the pressure at the end-systolic state while fixing
the remaining quantities were made. This estimate of end-systolic
elastance where also significantly higher in the healthy control
group.

\subsection{Paper 3: Mechanical Analysis of pulmonary hypertension via
  adjoint based data assimilation of a finite element model.}

\subsection{Paper 4: Assesment region regional myocardial work using
  adjoint-based data assimilation}

\newpage
\section{Other contributions}
Along with the research articles presented in this thesis, other
types of contributions in terms of talks, posters and software has
been made during the writing of this thesis. These contributions are
listed below.

@@

\subsection{Software}
\begin{itemize}
  \item Pulse-Adjoint, FEniCS-based cardiac mechanics solver and data
    assimilator, source: \url{https://bitbucket.org/finsberg/pulse_adjoint}
  \item Mesh-Toolbox, Toolbox for generating FEniCS meshes from 4D
    Echo,  source: \url{https://bitbucket.org/finsberg/mesh_generation}
\end{itemize}

\newpage
\section{Closing remarks and future directions}

Although we have shown in this thesis that
adjoint-based data assimilation is a powerful technique that opens new
possibilities in terms of patient specific mechanical simulations, many questions
still has to be answered before we can fully embrace the output of
such simulations.

First of all, is should be clear that \emph{the
  quality of biomarkers you can extract from a data-driven
  model cannot be any better than the data used as input to the
  model}. A typical saying is that garbage in $=$ garbage out,
meaning that if the data you use to constrain the model is noisy, then
you will also fit this noise if you allow for enough degree of
freedom. Regularization techniques (Section
\ref{sec:intro_regularization}) provides a way to attack this problem,
but it is not clear what is the best approach. If the noise in the
data is normally distributed with zero expected value, then adding
more data to the cost functional will also have a regularizing effect.
Hence, as the data assimilation techniques developed in this thesis
can take into account large amount of data, as much data as possible
should be provided to the assimilator.

A rule of thump is that \emph{the spatial resolution of the parameters should be
  reflected in the spatial resolution of the observations}. This means
that if one is trying to fit data that are spatially resolved at some level,
then choosing parameters that are resolved at a finer level should be
done with caution. Regarding both paper 1 and 2 we see that the
spatial resolution chosen was at a much finer level than the input
data. In this case, regularization techniques was used to restrict the
parameter space.

When estimating high dimensional parameters, a natural question
concerning uniqueness of these estimates arises. It is obvious that if one allows for
enough degree of freedom in the parameter space, then it is possible
to fit almost any type of data. Therefore \emph{more work on ensuring
identifiability} of these estimates (Section
\ref{sec:intro_identifiability}) should be done. If the output of such
model should have any clinical utility, then uniqueness of the
estimated parameters absolutely pivotal. Moreover, \emph{validation} of
these model is what matter most in terms of translating such computational
models into the clinic.

Regarding the mechanical modeling of the heart, 
\emph{choosing appropriate boundary conditions} that reflect the reality has
been a problem during the work in this thesis, and several different
choices has been made. Moreover, \emph{accounting for the orthotropic as
well as the visco-elastic behavior of the myocardium} is something that should
be investigated in future studies. Accounting for an orthotropic
behavior would acquire more parameters to be estimated, which would
thus require more input data.

In this thesis we have also not fully
explored the \emph{spatial resolution of the material parameters}, and it
should be investigated whether it is possible to relate locally estimated
tissue stiffness to e.g myocardial infarction.

The choice of active model for the myocardium is another topic that
should be addressed in more detail in future studies. In the author's
opinion, the output of the two fundamentally different approaches may
vary a lot. In particular, it is evident that the active strain
formulation do a better job in fitting strain data. One
hypothesis is that the amount of transverse active stresses should be
adjusted to each individual. These transverse active stresses are
naturally embedded in the active strain approach via a volume
preserving active deformation.

Finally, coupling the electrophysiology to the mechanical model within
this adjoint based data assimilation framework is something that
should also be investigated in the future.

% Here we start by listing a couple of statements
% which should be taken into considerations.

% \begin{itemize}
% \item \emph{The quality of features/biomarkers you can extract from a data-driven
% model cannot be any better than the data used as input to the
% model}. Most of the results presented in this thesis are based on data
% obtained from clinical measurements of real patients.
% \item \emph{The spatial resolution of the parameters should be
%     reflected in the spatial resolution of the observations.} When
%   trying to fit data that are spatially resolved at some level,
%   choosing parameters that are resolved at a finer level should be
%   done with caution. The continuity in the underlying physics as well
%   as regularization techniques could be used to ...

% \end{itemize}

% During the work of this thesis, several questions still remains open
% and would require

% \begin{itemize}
% \item The active model for the myocardium is has .. Degree of
%   tranverse activation. Matching of strain data..
% \item Identifiability of parameters... Uniqueness of
%   solutions.. Amount of regularization.. Convexity of the mismatch functional
% \item Appropriate boundary conditions.. In paper three we saw big
%   differences in the choice of boundary conditions. Especially, the
%   magnitude of the stress seems to 
% \item Coupling of electrophysiologigy and mechanics in an
%   ajoint-based. data assimilation framework. 
% \end{itemize}
